Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-26 paper/main.tex
content presentable-ish

@@

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 \citep{howe2020ensembl} as well as all non-embargoed plant
species from the JGI Phytosome 13 database \citep{goodstein2012phytozome}; exact 
genomes used are list here \ref{suptab:source_data}. % add a supplemental table with names and versions
Data was downloaded on 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

Both to divide genomes into training, validation and test sets, as well as to understand
performance, we needed a way to evaluate genome and annotation quality.
Metadata was generated for each genome, including: basic length and 
composition statistics \citep{gurevich2013quast}; k-mer content \citep{marccais2011fast};
conserved gene content of the genome, transcriptome, and proteome (\cite{simao2015busco};
file conversion with gffread from \cite{trapnell2012differential}; 
and finally simple counts of genic features from the gff files. In particular, 
we considered the following features of the metadata when judging the quality of a genome and it's 
annotation: assembly contiguity (N50 / genome size); retention in the proteome 
of detected conserved orthogroups (BUSCOs) that were identified in the genome; and
the presence of UTRs and alternative splicing as estimated by the gff feature-counts.
Finally this metadata was combined with researcher knowledge of model species to guide genome
selection as necessary.

At the start of our work we split both genome groups into {\it training genomes} and 
{\it test genomes}. The test genome groups are made up of 19 and 6 genomes, 
respectively, and were chosen to be of decent quality and diversity based on collected 
metadata while also representing a broad phylogenetic spread.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with "new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with "optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between AUGUSTUS and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

@@

We include trained models for plants and animals which achieve 
high prediction accuracy on cross-species gene annotation for broad 
phylogenetic groups (land plants and vertebrates, respectively).
Within these groups, this eliminates the dependency on retraining and the
expertise and data required therefore. Production of comparable, single-method
annotations for broad groups has the potential to greatly facilitate downstream
analyses which suffer when data is heterogeneous \citep{}.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables
\begin{suptab}\label{suptab:rnaseq_samples}\end{suptab}
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:rna_select_animals}\end{supfig}
\begin{supfig}\label{supfig:rna_select_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_03}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{ssec:data_prep}\end{supsec}
\begin{supsec}\label{ssec:generalization}\end{supsec}
\begin{supsec}\label{ssec:hyperparameters}\end{supsec}
\begin{supsec}\label{ssec:longer}\end{supsec}
\begin{supsec}\label{ssec:overlapping}\end{supsec}
\begin{supsec}\label{ssec:training_data}\end{supsec}
\begin{supsec}\label{ssec:dcnn}\end{supsec}
\begin{supsec}\label{ssec:rnaseq_eval}\end{supsec}

>>>>

\newcounter{supdata}
\newenvironment{supdata}[1][]{\refstepcounter{supdata}}

@@

\begin{methods}
\section{Methods}
\subsection{Datasets}
The foundation of our work are 192 animal and 60 plant genomes. The data of each 
genome consists of the latest publically available genomic assembly in form of a 
FASTA file and the latest annotation in the GFF format. One genome for each species of animal 
in EnsemblMetazoa 45 (\cite{howe2020ensembl}; Table S\ref{suptab:downloads_animals}) 
as well as all non-embargoed plant species from the JGI Phytosome 13 database 
(\cite{goodstein2012phytozome}; Table S\ref{suptab:downloads_plants}); exact 
genomes used are list here . % add a supplemental table with names and versions
Data was downloaded on 15/10/2019 and 29/03/2019 for animals and plants respectively. 
Both data groups were used seperately during training as well as during evaluation.

At the start of our work we split both genome groups into {\it training genomes} and 
{\it test genomes}. The test genome groups are made up of 19 and 6 genomes, 
respectively, and were chosen to be of decent quality and diversity based on collected 
metadata (see section S\ref{ssec:data_prep} and Dataset S\ref{supdata:metadata}) 
while also representing a broad phylogenetic spread.

@@

Orthologs of highly conserved generally single-copy genes were identified in each genome using
BUSCO. The {\it viridiplantae} set was used for plants and the {\it metazoa} set for animals.
The training genbank files were generated directly by BUSCO, by utilizing the ``--long"  % inline code?
parameter. For plants, the entire retraining could be performed as above; however, for animals
we randomly selected only half the BUSCO-generated training set, which resulted in a training set 
size and runtime comparable to the plants (~400 genes and several days per species). For animals
the training was carried out with the subsetted training file and using the following scripts provided
by AUGUSTUS. An untrained model was setup with ``new\_species.pl", and the model was fit by running
"etraining" before and after the major hyper-parameter optimization with ``optimize\_augustus.pl".
Using the trained model for each species, we ran the main prediction ("augustus") with ``--UTR=off"
and ``--gff3=on". The gffs produced by AUGUSTUS were imported into GeenuFF and exported as h5
files in the same manor as the reference, allowing direct comparison with reference and 
Helixer predictions. While tenable for some 237 species, this method allowed neither training
nor prediction of UTR regions with AUGUSTUS, so the metric Subgenic F1, which excludes UTR,
was used for all comparisons between AUGUSTUS and Helixer.

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
AUGUSTUS or HMMs as AUGUSTUS was used as the {\it de novo} gene caller for the reference of
many plant species % (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
%{\it Asparagus officinalis}, 
%\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more. %(e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
%{\it Capsella grandiflora}, \cite{slotte2013capsella}). 
The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

@@

We include trained models for plants and animals which achieve 
high prediction accuracy on cross-species gene annotation for broad 
phylogenetic groups (land plants and vertebrates, respectively).
Within these groups, this eliminates the dependency on retraining and the
expertise and data required therefore. Production of comparable, single-method
annotations for broad groups has the potential to greatly facilitate downstream
analyses; for instance it could avoid some of the inconsistency and errors that
are otherwise seen in RNAseq analyses when different annotations are used 
\citep{zhao2015comprehensive,torres2016robust}.

@@

\bibliographystyle{natbib}
\bibliography{literature}
% place holders for supplemental things
% ori version was grepped out of supplements, then
% :'<,'>s/^/\\begin{supfig}/g
% :'<,'>s/$/\\end{supfig}/g
% and similar for tables/sections
% tables

\begin{suptab}\label{suptab:downloads_animals}\end{suptab}
\begin{suptab}\label{suptab:downloads_plants}\end{suptab}
\begin{suptab}\label{suptab:generalization_animals}\end{suptab}
\begin{suptab}\label{suptab:generalization_plants}\end{suptab}
\begin{suptab}\label{suptab:params}\end{suptab}
\begin{suptab}\label{suptab:classweights}\end{suptab}
\begin{suptab}\label{suptab:rnaseq_samples}\end{suptab}
\begin{suptab}\label{suptab:prediction_lengths}\end{suptab}
\begin{suptab}\label{suptab:genome_stats}\end{suptab}
\begin{suptab}\label{suptab:dcnn_xval}\end{suptab}
\begin{suptab}\label{suptab:dcnn_params}\end{suptab}
% figures
\begin{supfig}\label{supfig:phylogeny_f1_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_animals}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_f1_plants}\end{supfig}
\begin{supfig}\label{supfig:phylogeny_differences_plants}\end{supfig}
\begin{supfig}\label{supfig:rna_select_animals}\end{supfig}
\begin{supfig}\label{supfig:rna_select_plants}\end{supfig}
\begin{supfig}\label{supfig:overlapping_animals}\end{supfig}
\begin{supfig}\label{supfig:overlapping_plants}\end{supfig}
\begin{supfig}\label{supfig:cov_example_02}\end{supfig}
\begin{supfig}\label{supfig:cov_example_03}\end{supfig}
\begin{supfig}\label{supfig:UTRs_animals}\end{supfig}
\begin{supfig}\label{supfig:UTRs_plants}\end{supfig}
\begin{supfig}\label{supfig:average_vs_ref}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_augustus}\end{supfig}
\begin{supfig}\label{supfig:each_sp_vs_ref}\end{supfig}
% sections
\begin{supsec}\label{ssec:data_prep}\end{supsec}
\begin{supsec}\label{ssec:generalization}\end{supsec}
\begin{supsec}\label{ssec:hyperparameters}\end{supsec}
\begin{supsec}\label{ssec:rnaseq_setup}\end{supsec}
\begin{supsec}\label{ssec:longer}\end{supsec}
\begin{supsec}\label{ssec:overlapping}\end{supsec}
\begin{supsec}\label{ssec:training_data}\end{supsec}
\begin{supsec}\label{ssec:dcnn}\end{supsec}
\begin{supsec}\label{ssec:rnaseq_eval}\end{supsec}
% datasets
\begin{supdata}\label{supdata:metadata}\end{supdata}  % just a placeholder for the CSV and where it's referenced
